---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# R Iterator Ideas (`[[` + signal, optionally with `as.iterable()`)

This approach extends the behavior of `[[` to signal a custom condition
when done:

```{r}
stop_out_of_bounds <- function(call = sys.call()) {
  cnd <- structure(
    list(message = "subscript out of bounds", call = call),
    class = c("out_of_bounds", "error", "condition")
  )
  stop(cnd)
}
```

Which we use from `[[`:

```{r, error = TRUE}
# Quick and dirty implementation; real implementation would need to happen in C
`[[` <- function(x, i, ...) {
  if (is.vector(x) && is.integer(i) && length(i) == 1 && i > length(x)) {
    stop_out_of_bounds()
  } else {
    base::`[[`(x, i, ...)
  }
}
x <- 1:3
x[[5L]]
```

We additionally need a protocol to turn an object into an iterator, so,
if needed, `[[`, can behave differently outside of `for`:

```{r}
as.iterable <- function(x) UseMethod("as.iterable")
as.iterable.default <- function(x) x
```

The `for` repeatedly increments i and calls `x[[i]]`:

```{r}
`for` <- function(var, iterable, body) {
  var <- as.character(substitute(var))
  body <- substitute(body)
  env <- parent.frame()

  if (!is.vector(iterable) || is.object(iterable))
    iterable <- as.iterable(iterable)
  
  i <- 1L
  exhausted <- FALSE
  repeat {
    tryCatch(
      value <- iterable[[i]],
      out_of_bounds = function(cond) {
        exhausted <<- TRUE
      }
    )
    if (exhausted) return(invisible())
    
    assign(x = var, value = value, envir = env)
    eval(body, env)
    
    i <- i + 1L
  }
}

for (x in 1:5) {
  str(x)
}
```

### Calculation on demand

```{r}
SquaresSequence <- function(from = 1, to = 10) {
  structure(
    list(from = from, to = to),
    class = "SquaresSequence"
  )
}

`[[.SquaresSequence` <- function(x, i) {
  val <- x$from - 1L + i
  if (val > x$to) {
    stop_out_of_bounds()
  } else {
    val ^ 2  
  }
}

for (x in SquaresSequence(5, 10))
  print(x)
```

## Sequence of unknown length

Handling a sequence of unknown length is a bit tougher because we need
to create an intermediate mutable object that tracks state.

```{r}
SampleSequence <- function(max) {
  structure(
    list(max = max),
    class = "SampleSequence"
  )
}

as.iterable.SampleSequence <- function(x) {
  state <- new.env(parent = emptyenv())
  state$total <- 0
  
  structure(
    list(max = x$max, state = state),
    class = "SampleSequenceIterator"
  )
}

`[[.SampleSequenceIterator` <- function(x, i) {
  val <- abs(rnorm(1))
  x$state$total <- x$state$total + val
  
  if (x$state$total > x$max) {
    stop_out_of_bounds()
  } else {
    val
  }
}

for (x in SampleSequence(2)) {
  print(x)
}
```

## Extensions

### Iterators

```{r}
as.iterable.function <- function(x) {
  # Not strictly necessary, just a guardrail here so we wouldn't have to part
  # with our old friend: "object of type 'closure' is not subsettable"
  times_called <- 0L
  iterator <- function() {
    on.exit(times_called <<- times_called + 1L)
    x()
  }
  class(iterator) <- "iterable_function"
  iterator
}

`[[.iterable_function` <- function(x, i) {
  if(environment(x)$times_called + 1L != i)
    warning("Something is wrong")
  x() # x is expected to eventually call stop(signal_out_of_bounds())
}
```

```{r}
SquaresSequenceClosure <- function(from = 1, to = 10) {
  force(from); force(to)
  function() {
    from <<- from + 1L
    if(from > to)
      stop_out_of_bounds()
    from ^ 2
  }
}

for (x in SquaresSequenceClosure(5, 10))
  print(x)
```

### Reticulate

Reticulate support might look like this:

```{r}
as.iterable.python.builtin.object <- function(x) {
  structure(
    list(it = reticulate::as_iterator(x)),
    class = "reticulate_iterable"
  )
}

`[[.reticulate_iterable` <- function(x, i) {
  sentinel <- environment()
  val <- reticulate::iter_next(x$it, completed = sentinel)
  if(identical(val, sentinel))
    stop_out_of_bounds()
  val
}

for(x in reticulate::r_to_py(1:3))
  print(x)
```
