---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# R Iterator Ideas

R doesn't currently have a mechanism for users to control what happens
when objects are passed to `for`. Below are two different sketches for
what such a mechanism might look like:

1.  User implements `[[` , and optionally `as.iterable()` . User is
    expected to call `stop(out_of_bounds_condition())` in `[[` when the
    iterator is exhausted. `for` repeatedly calls `x[[i]]` with auto
    incrementing `i` until the condition is encountered.

2.  User implements `iterate()`, which is called with arguments
    `iterate(x, state)`. `iterate()` is expected to return a list of
    length 2: `list(next_elem, next_state)`. `for` repeatedly calls
    `iterate(x, state)` until `next_state` is `NULL`.

All approaches sketched out would also accept a regular R function in
`for`, with the function repeatedly called until a signal is raised.

### `[[` + signal, optionally with `as.iterable()`

The equivalent of the following could be in base R (with core parts
implemented in C):

```{r}
as.iterable <- function(x) UseMethod("as.iterable")
as.iterable.default <- identity

`for` <- function(var, iterable, body) {
  var <- as.character(substitute(var))
  body <- substitute(body)
  env <- parent.frame()

  iterable <- as.iterable(iterable)
  i <- 1L
  exhausted <- FALSE
  repeat {
    tryCatch({
      value <- iterable[[i]]
    },
    out_of_bounds = function(cond) {
      exhausted <<- TRUE
    },
    error = function(e) {
      # Not everything raises a typed signal yet, so accept simple errors too
      if (e$message == "subscript out of bounds")
        exhausted <<- TRUE
      else
        stop(e)
    })
    if (exhausted) return(invisible())
    assign(x = var, value = value, envir = env)
    i <- i + 1L
    eval(body, env)
  }
}

out_of_bounds_condition <- function(call = sys.call()) {
  structure(class = c("out_of_bounds", "error", "condition"),
            list(message = "subscript out of bounds", call = call))
}

# support for functions
as.iterable.function <- function(x) {
  # Not strictly necessary, just a guardrail here so we wouldn't have to part
  # with our old friend: "object of type 'closure' is not subsettable"
  times_called <- 0L
  iterator <- function() {
    on.exit(times_called <<- times_called + 1L)
    x()
  }
  class(iterator) <- "iterable_function"
  iterator
}

`[[.iterable_function` <- function(x, i) {
  if(environment(x)$times_called + 1L != i)
    warning("Something is wrong:")
  x() # x is expected to eventually call stop(signal_out_of_bounds())
}
```

With this approach, the users might write code like this:

```{r}
SquaresSequence <- function(from = 1, to = 10) {
  out <- list(from = from, to = to)
  class(out) <- "SquaresSequence"
  out
}

`[[.SquaresSequence` <- function(x, i) {
  if (i > x$to)
    stop(out_of_bounds_condition())
  (x$from - 1L + i) ^ 2
}

for (x in SquaresSequence(1, 10))
  print(x)


SquaresSequenceClosure <- function(from = 1, to = 10) {
  force(from); force(to)
  function() {
    from <<- from + 1L
    if(from > to)
      stop(out_of_bounds_condition())
    from ^ 2
  }
}

for (x in SquaresSequenceClosure(1, 10))
  print(x)
```

Reticulate support might look like:

```{r}
as.iterable.python.builtin.object <- function(x) {
  iterator <- reticulate::as_iterator(x)
  class(iterator) <- unique(c("reticulate_iterator", class(iterator)))
  iterator
}

`[[.reticulate_iterator` <- function(x, i) {
  sentinal <- environment()
  val <- reticulate::iter_next(x, completed = sentinal)
  if(identical(val, sentinal))
    stop(out_of_bounds_condition())
  val
}

for(x in reticulate::r_to_py(1:3))
  print(x)
```

```{r, include = FALSE}
rm(list = ls(all.names = TRUE)) # see no evil :P
```

### `iterate(x, state)` (The Julia approach)

The equivalent of the following could be in base R (with core parts
implemented in C):

```{r}
envir::import_from(zeallot, `%<-%`) # tuple unpacking assignment operator

iterate <- function(x, state = NULL) UseMethod("iterate")
iterate.default <- function(x, state = NULL) {
  if(is.null(state))
    state <- 1L
  if(state > length(x))
    return(list(NULL, NULL))
  list(x[[state]], state + 1L)
}

`for` <- function(var, iterable, body) {
  var <- as.character(substitute(var))
  body <- substitute(body)
  env <- parent.frame()

  state <- NULL
  repeat {
    c(value, state) %<-% iterate(iterable, state)
    if(is.null(state)) return(invisible())
    assign(x = var, value = value, envir = env)
    eval(body, env)
  }
}


iterate.function <- function(x, state = NULL) {
  if(is.null(state))
    state <- x
  value <- NULL
  tryCatch(value <- state(),
           iterator_exhausted = function(e) state <<- NULL)
  list(value, state)
}


iterator_exhausted_condition <- function(call = sys.call()) {
  structure(class = c("iterator_exhausted", "error", "condition"),
            list(message = "iterator exhausted", call = call))
}
```

With this approach, the users might write code like this:

```{r}
SquaresSequence <- function(from = 1, to = 10) {
  out <- list(from = from, to = to)
  class(out) <- "SquaresSequence"
  out
}

iterate.SquaresSequence <- function(x, state = NULL) {
  if(is.null(state))
    state <- 1L
  if (state > x$to)
    return(list(NULL, NULL))
  list((x$from - 1L + state) ^ 2, state + 1L)
}


for (x in SquaresSequence(1, 10))
  print(x)


SquaresSequenceClosure <- function(from = 1, to = 10) {
  force(from); force(to)
  function() {
    from <<- from + 1L
    if(from > to)
      stop(iterator_exhausted_condition())
    from ^ 2
  }
}

for (x in SquaresSequenceClosure(1, 10))
  print(x)
```

Reticulate support might look like

```{r}
iterate.python.builtin.object <- function(x, state = NULL) {
  if(is.null(state))
    state <- reticulate::as_iterator(x)

  sentinal <- environment()
  value <- reticulate::iter_next(state, completed = sentinal)
  if(identical(sentinal, value))
    list(NULL, NULL)
  else
    list(value, state)
}

for(x in reticulate::r_to_py(1:3))
  print(x)
```
